# Adjustment {#sec-adjustment}

```{r label='250-Adjustment-4QgjSm', include=FALSE}
source("../_startup.R")
set_chapter(11)
```

The phrase "all other things being equal" is a critical qualifier in describing relationships.  To illustrate: A simple claim in economics is that a high price for a commodity reduces the demand. For example, increasing the price of gasoline will reduce demand as people avoid unnecessary driving or purchase electric cars. Nevertheless, the claim can be considered obvious only with the qualifier *all other things being equal*.  For instance, the fuel price might have increased because a holiday weekend and the attendant vacation travel has increased the demand for gasoline.  Thus, higher gasoline prices may be associated with higher demand unless holding constant other variables such as vacationing.  

The Latin equivalent of "all other things being equal" is sometimes used in economics: "**ceteris paribus**". The economics claim would be, "higher prices are associated with lower demand, *ceteris paribus*." 

Although the phrase "all other things being equal" has a logical simplicity, it is impractical to implement "all." So instead of the blanket "all other things," it is helpful to consider just "some other things" to be held constant, being explicit about what those things are. Other phrases along the same lines are "adjusting for ...,"  "taking into account ...," and "controlling for ...."  

## Groupwise adjustment

"**Life expectancy**" is a statistical summary familiar to many readers. Life expectancy is often the evidence provided in debates about healthcare policies or environmental conditions. For instance, consider this pull-quote from the [*Our World in Data* website](https://ourworldindata.org/us-life-expectancy-low):

> "*Americans have a lower life expectancy than people in other rich countries despite paying much more for healthcare.*"


::: {#tbl-life-expectancy .column-margin .content-visible when-format="html"}
Country | Female | Male
--------|----|-----
Japan   | 87.6 | 84.5
Spain   | 86.2 | 80.3
Canada  | 84.7 | 80.6
United States | 80.9 | 76.0
Bolivia | 74.0 | 71.0
Russia  | 78.3 | 66.9
North Korea | 75.9 | 67.8
Haiti   | 68.7 | 63.3
Nigeria | 63.3 | 59.5
Somalia | 58.1 | 53.4

Life expectancy at birth for several countries and territories. [Source](https://en.wikipedia.org/wiki/List_of_countries_by_life_expectancy) 
:::

<!-- for PDF version -->
::: {.column-margin .content-visible when-format="pdf"}
Country | Female | Male
--------|----|-----
Japan   | 87.6 | 84.5
Spain   | 86.2 | 80.3
Canada  | 84.7 | 80.6
United States | 80.9 | 76.0
Bolivia | 74.0 | 71.0
Russia  | 78.3 | 66.9
North Korea | 75.9 | 67.8
Haiti   | 68.7 | 63.3
Nigeria | 63.3 | 59.5
Somalia | 58.1 | 53.4

table: Life expectancy at birth for several countries and territories. [Source](https://en.wikipedia.org/wiki/List_of_countries_by_life_expectancy) 
:::

The numbers in @tbl-life-expectancy faithfully reflect the overall situation in the different countries. Yet, without adjustment, they are not well suited to inform about specific situations. For example, life expectancies are usually calculated *separately* for males and females, acknowledging a significant association of life expectancy with sex, not just the availability of medical care. We will call such a strategy "**groupwise adjustment**" because it's based on acknowledging difference between groups. You'll see similar groupwise adjustment of life expectancy on the basis of race/ethnicity.

Over many years teaching epidemiology at Macalester College, I asked students to consider life-expectancy tables and make policy suggestions for improving things. Almost always, their primary recommendations involved improving access to health care, especially for the elderly.

But life expectancy is not mainly, or even mostly, about old age. Two critical determinants are infant mortality and lethal activities by males in their late teenage and early adult years. If we want to look at conditions in the elderly, we need to consider elderly people separately, not mixed in with infants, children, and adolescents. For reasons we won't explain here, with life expectancy calculations it's routine to calculate a separate "life expectancy at age X" for each age year.  @tbl-life-expectancy-at-70 shows, according to the World Health Organization, how many years longer a 70-year old can expect to live. The 30-year difference between Japan and Somalia seen in @tbl-life-expectancy is reduced, for 70-year olds, to about a decade. The differences between males and females are similarly reduced 

::: {#tbl-life-expectancy-at-70 .column-margin .content-visible when-format="html"}

Country | Female | Male
--------|-----|-----
Japan | 21.3 | 17.9 
Canada | 18.0 | 15.6 
Spain | 17.0 | 14.0
United States | 18.3 | 16.3
Russia | 16.2 | 12.2
Bolivia | 13.6 | 13.0
Haiti | 12.9 | 12.1
Somalia | 11.6 | 9.7

Life expectancy at age 70. (Main source: [World Health Organization](https://apps.who.int/gho/data/view.main.61780?lang=en)) average of 65-74 year olds) 
:::

<!-- PDF version of above -->
::: {.column-margin .content-visible when-format="pdf"}

Country | Female | Male
--------|-----|-----
Japan | 21.3 | 17.9 
Canada | 18.0 | 15.6 
Spain | 17.0 | 14.0
United States | 18.3 | 16.3
Russia | 16.2 | 12.2
Bolivia | 13.6 | 13.0
Haiti | 12.9 | 12.1
Somalia | 11.6 | 9.7

table: Life expectancy at age 70. (Main source: [World Health Organization](https://apps.who.int/gho/data/view.main.61780?lang=en)) average of 65-74 year olds) 
:::

## Adjustment with *per* {#sec-per-adjustment}

The US government's Centers for Medicare Studies gives [some numbers about the age distribution](https://www.cms.gov/Research-Statistics-Data-and-Systems/Statistics-Trends-and-Reports/NationalHealthExpendData/Downloads/AgeandGenderHighlights.pdf) of "personal health-care" spending: 

> "*In 2020, children (0-18) accounted for 23 percent of the population and 10 percent of personal health care (PHC) spending, working age adults (19-64) accounted for 60 percent of the population and 53 percent of PHC, and older adults (65 and older) account for 17 percent of the population and 37 percent of PHC.*"

::: {.column-margin .content-visible when-format="pdf"}
group | age span | population | PHC spending 
:--------|----:|------------|----------
children | 0-18      | 23%        | 10%  
working age | 19-64     | 60%        | 53%
older adults | 65+       | 17%        | 37%

Table 3: A tabular arrangement of the data from the Centers for Medicare Studies
:::

::: {#tbl-PHC-spending .column-margin .content-visible when-format="html"}
group | age span | population | PHC spending 
:--------|----:|------------|----------
children | 0-18      | 23%        | 10%  
working age adults| 19-64     | 60%        | 53%
older adults | 65+       | 17%        | 37%

A tabular arrangement of the data from the Centers for Medicare Studies
:::

The textual presentation of data presents relevant information but obscures the patterns. The author would have done better by placing the numbers that are to be compared to one another next to one another. A tabular organization makes it much easier to compare the relative population sizes of the age groups. For instance, the population column of @tbl-PHC-spending shows at a glance that the population of children and older adults are about the same.





Similarly, the table's "PHC spending" column makes it obvious that PHC spending is much higher for working age adults than for either children or older adults.

In comparing the spending between groups, it can be helpful to *take into account* the differing population sizes. A **per capita** adjustment---spending divided by population size---accomplishes it. For instance, the per capita adjustment for children is 10% / 23%, that is, 0.43.  @tbl-per-capita-spending shows the *per capita* spending for all three age groups.

::: {#tbl-per-capita-spending}
group | age | population | spending | spending per capita
:--------|----:|------------|----------|---------
children | 0-18      | 23%        | 10%      | 0.43
working age adults| 19-64     | 60%        | 53%      | 0.88
older adults | 65+       | 17%        | 37%      | 2.18

Adjusting spending for the size of the population gives a clearer indication of how spending compares between the different age groups.
:::

Including the *per capita* adjusted spending in the table makes it easy to see an important pattern: older adults have much higher health spending (per person) than the other groups.

The literal meaning of "*per capita*" is "for each head." But the method of adjusting by dividing one quantity by another has much broader applications. To illustrate, let's return to the example of college grades from @sec-grade-joins. There, we calculated using simple wrangling each student's grade-point average and an instructor grade-giving average. The instructor's grade-giving average varies so much that it seems short-sighted to neglect it as a factor in determining a student's grade in that instructor's courses.

An adjustment for the instructor can be made by constructing a *per*-type index. An instructor gave each grade, but instead of considering the grade literally, let's divide the grade by the grade-giving average of the instructor involved.

::: {.column-margin .content-visible when-format="pdf"}
  sid       iid      sessionID    gradepoint 
-------- --------- ------------- ------------
 S32418   inst268   session2911       3 
 S32328   inst436   session3524      3.66
 S32250   inst268   session2911      2.66
 S32049   inst436   session2044      3.33
 S31914   inst436   session2044      3.66
 S31905   inst436   session2044       4 
 S31833   inst436   session3524      3.33
 S31461   inst264   session1904       4 
 S31197   inst436   session3524       3 
 S31194   inst264   session1904       2

A few of the 6,124 rows in the `Extended_grades` table.
:::

We can consider the instructors' `iGPA` to calculate an instructor-adjusted GPA for students. We create a data frame with the instructor ID and numerical grade point for every grade in the `Grades` and `Sessions` tables. First, we use "joins" to bring together the tables from the database. 




```{r message = FALSE}
Extended_grades <- Grades |> 
  left_join(Sessions) |>
  left_join(Gradepoint) |>
  select(sid, iid, sessionID, gradepoint)
```

```{r echo=FALSE}
# for printing purposes
set.seed(204) 
Foobar <- Extended_grades |> 
  filter(sessionID %in% 
           c("session2911", "session1904", "session3524", "session2491",
             "session3822", "session2044", "session2606")) |>
  sample(n = 4, .by = sid) |> unique() |>
  sample(n=2, .by = gradepoint) |> arrange(desc(sid)) |>
  filter(!is.na(gradepoint))
```

::: {#fig-ex-grades .column-margin .content-visible when-format="html"}
```{r and_so_on="... for 6,124 rows altogether", echo = FALSE}
Foobar
# set.seed(109)
# Who_to_show <- Extended_grades |>
#   filter(sid %in% !!head(Foobar$sid, 8)) |>
#   sample(n = 10) 
# Who_to_show 
```
:::





Next, calculate the instructor-by-instructor "grade-giving average" (`gga`):

```{r}
Instructors <- Extended_grades |>
  summarize(gga = mean(gradepoint, na.rm = TRUE), .by = iid)
```

```{r echo = FALSE}
Show_these <- Instructors |> 
  filter(iid %in% !!Foobar$iid)
```

::: {.column-margin .content-visible when-format="html"}
```{r}
Show_these
```

Three rows from the `Instructors` data frame.
:::

::: {.column-margin .content-visible when-format="pdf"}
   iid      gga  
--------- -------
 inst436   3.584 
 inst264   2.974 
 inst268   3.062 

Three rows from the `Instructors` data frame.
:::


Joining the `Instructors` data frame with `Extended_grades` puts the grade earned and the average grade given next to one another. The unit of observation is still a student receiving a grade in a class session.

```{r message = FALSE}
With_instructors <- 
  Extended_grades |>
  left_join(Instructors)
```

```{r label='250-Adjustment-1vQRlK', echo=FALSE, and_so_on = "... for 364 instructors altogether", digits=3}
set.seed(1091)
Display2 <- With_instructors |> 
  filter(iid %in% !!Foobar$iid) |>
  sample(2, .by = iid)
```

::: {.content-visible when-format="pdf"}
  sid       iid      sessionID    gradepoint    gga  
-------- --------- ------------- ------------ -------
 S32310   inst436   session2193      3.66      3.584 
 S31794   inst436   session2541      3.66      3.584 
 S32289   inst264   session2235       4        2.974 
 S31461   inst264   session1904       4        2.974 
 S32211   inst268   session2650      2.33      3.062 
 S32250   inst268   session2911      2.66      3.062 
 
A few rows of he result of joining `Extended_grades` with `Instructors`. The data frame has 6,124 rows in total.
:::

::: {.content-visible when-format="html"}
```{r echo=FALSE, and_so_on = "... for 6,124 rows altogether", digits=3}
Display2
```

Joining `Extended_grades` with `Instructors`.
:::

Make the *per* adjustment by dividing `gradepoint` by `gga` to create a grade index. We will then average this index for each student to create each student's instructor-adjusted GPA (`adj_gpa`), shown in @tbl-adj-gpa.

```{r}
Adjusted_gpa <-
  With_instructors |>
  mutate(index = gradepoint / gga) |>
  summarize(adj_gpa = mean(index, na.rm = TRUE), .by = sid)
```

```{r echo=FALSE}
set.seed(109)
Display3 <- Adjusted_gpa |> 
  filter(sid %in% !!Foobar$sid) |>
  sample(8)
```

::: {#tbl-adj-gpa .column-margin .content-visible when-format="html"}
```{r echo=FALSE, and_so_on = "... for 443 students altogether.", digits=3}
Display3
```
:::

::: {.column-margin .content-visible when-format="pdf"}
  sid     adj_gpa 
-------- ---------
 S31197   0.9578  
 S31914    1.04   
 S31461    1.148  
 S32250   0.9284  
 S31194   0.9978  
 S32418   0.9328  

... for 443 rows altogether
:::


## Adjustment by modeling {#sec-adjustment-by-modeling}

We will use the word "**adjustment**" to name the statistical techniques by which "other things" are considered. Those other things, as they appear in data, are called "**covariates**."

There are two phases for modelling-based adjustment, one requiring careful thought and understanding of the specific system under study, the other---the topic of this section---involving only routine, straightforward modeling calculations. 

**Phase 1**: Choose relevant covariates for adjustment. This almost always involves familiarity with the real-world context. 

**Phase 2**: Build a model with the covariates from Phase 1 as explanatory variables. 

To illustrate, we return to the college grades example in @sec-per-adjustment. There, we did a *per* adjustment of each grade by the average of all the grades assigned by the instructor (the "grade-giving average": `gga`). 

Now we want to examine how to incorporate other factors into the adjustment, for instance class size (`enroll`) and class `level`. We will also change from the politically unpalatable instructor-based grade-given average to using department (`dept`) as a covariate.

To start, we point out that the conventional GPA can also be found by modeling `gradepoint ~ sid`. 

```{r label='250-Adjustment-gWgaoo', message = FALSE}
Joined_data <-   Grades |> 
  left_join(Sessions) |>
  left_join(Gradepoint) 
Raw_model <- 
  Joined_data |> 
  model_train(gradepoint ~ sid)
```

The model values from `Raw_model` will be the unadjusted (raw) GPA. We can compute those model values by making a data frame with all the input values for which we want an output:

```{r}
Students <- Grades |> select(sid) |> unique()
```

Now evaluate `Raw_model` for each of the inputs in `Students` to find the model value (called `.output` by `model_eval()`).

```{r}
Raw_gpa <- Raw_model |>
  model_eval(Students) |>
  select(sid, raw_gpa = .output)
```

The advantage of such a modeling approach is that we can add covariates to the model specification in order to adjust for them. To illustrate, we will adjust using `enroll`, `level`, and `dept`:

```{r label='250-Adjustment-N2ieFI'}
Adjustment_model <-
  Joined_data |>
  model_train(gradepoint ~ sid + enroll + level + dept)
```  

As we did before with `Raw_model`, we will evaluate `Adjustment_model` at all values of `sid`. But we will also *hold constant* the enrollment, level, and department by setting their values. For instance, in the following, we look at every student as if their classes were all in department D, at the 200 level, and with an enrollment of 20.

```{r label='250-Adjustment-EgEazR'}
Inputs <- Students |>
  mutate(dept = "D", level = 200, enroll = 20)
Model_adjusted_gpa <-
  Adjustment_model |>
  model_eval(Inputs) |>
  rename(modeled_gpa = .output)
```

This is the core of adjustment: comparing individual specimens *after* putting them on the same footing, that is, *ceteris paribus*.

In @sec-adjustment-by-modeling,
we calculated three different versions of the GPA:

```{r echo=FALSE, message=FALSE}
# Save these so they can be used in an exercise.
Three_gpas <- Raw_gpa |>
  left_join(Adjusted_gpa) |>
  left_join(Model_adjusted_gpa)
save(Three_gpas, file = "www/Three-gpas.rda")
```

1. The *raw* GPA, which we calculated in two equivalent ways, with `summarize(mean(gradepoint), .by = sid)` and with the model `gradepoint ~ sid`.
2. The grade-given average used to create an index that involves `gradepoint / gga`.
3. The model using covariates `level`, `enroll`, and `dept`.

The statistical thinker knows that GPA is a social construction, not a hard-and-fast reality. Let's see to what extent the different versions agree.

