---
title: "Lessons in Statistical Thinking"
subtitle: "Intro Stats for the 21^st^ Century"
author: "Daniel T. Kaplan"
date: "Experimental PDF version: 2024-03-23"
format:
    pdf: 
        toc: true
        number-sections: true
        colorlinks: true
        shift-heading-level-by: 0
        crossref:
            chapters: true
        keep-tex: true
    html: 
      self-contained: true
      grid: 
        margin-width: 350px
reference-location: margin
citation-location: margin
---

```{r include=FALSE}
already_works <- TRUE
```

\newpage

```{r eval=already_works, child="L01-Data-frames.qmd"}
```

\newpage

```{r eval=already_works, child="L02-Pointplots.qmd"}
```

\newpage

```{r eval=already_works, child="L03-Variation-and-distribution.qmd"}
```

\newpage

```{r eval=already_works, child="L04-Model-annotations.qmd"}
```

\newpage

```{r eval=already_works, child="L05-Wrangling.qmd"}
```

\newpage

```{r eval=TRUE, child="L06-Computing.qmd"}
```

\newpage

```{r eval=TRUE, child="L07-Databases.qmd"}
```

\newpage

```{r eval=TRUE, child="L08-Statistical-thinking.qmd"}
```

\newpage

```{r eval=TRUE, child="L09-Accounting-for-variation.qmd"}
```

\newpage

```{r eval=TRUE, child="L10-Model-patterns.qmd"}
```

\newpage


```{r eval=TRUE, child="L11-Regression.qmd"}
```

\newpage

```{r eval=TRUE, child="L12-Adjustment.qmd"}
```
\newpage

```{r eval=TRUE, child="L13-Signal-and-noise.qmd"}
```

\newpage

```{r eval=already_works, child="L14-Simulation.qmd"}
```

\newpage

```{r eval=already_works, child="L15-Noise-patterns.qmd"}
```

\newpage

```{r eval=already_works, child="L16-Estimation-and-likelihood.qmd"}
```

\newpage

```{r eval=already_works, child="L17-R-squared.qmd"}
```

\newpage

```{r eval=already_works, child="L18-Prediction.qmd"}
```

\newpage

```{r eval=already_works, child="L19-Sampling-variation.qmd"}
```

\newpage

```{r eval=already_works, child="L20-Confidence-intervals.qmd"}
```

\newpage

```{r eval=already_works, child="L21-Measuring-and-accumulating-risk.qmd"}
```

\newpage

```{r eval=already_works, child="L22-Effect-size.qmd"}
```

\newpage

```{r eval=already_works, child="L23-DAGs.qmd"}
```

\newpage

```{r eval=already_works, child="L24-Causality-and-DAGs.qmd"}
```

\newpage

```{r eval=already_works, child="L25-Confounding.qmd"}
```

\newpage

```{r eval=already_works, child="L26-Experiment.qmd"}
```

\newpage

```{r eval=already_works, child="L27-Hypothetical-thinking.qmd"}
```

\newpage

\newcommand{\Ptest}{{\mathbb{P}}}
\newcommand{\Ntest}{{\mathbb{N}}}
\newcommand{\Sick}{{\cal S}}
\newcommand{\Healthy}{{\cal H}}
\newcommand{\given}{\ |\!\!|\  }

```{r eval=already_works, child="L28-Bayes.qmd"}
```

\newpage

```{r eval=already_works, child="L29-NHT.qmd"}
```

