<!-- Topic12-02 -->
::: {.callout-note collapse="true" #enr-12-02}
## DRAFT Arguing about adjustment



In @sec-adjustment-by-modeling,
we calculated three different versions of the GPA:

```{r label='Topic12-01-g3IXus', echo=FALSE, message=FALSE}
load("www/Three-gpas.rda")

```

1. The *raw* GPA, which we calculated in two equivalent ways, with `summarize(mean(gradepoint), .by = sid)` and with the model `gradepoint ~ sid`.
2. The grade-given average used to create an index that involves `gradepoint / gga`.
3. The model using covariates `level`, `enroll`, and `dept`.

The statistical thinker knows that GPA is a social construction, not a hard-and-fast reality. Let's see to what extent the different versions agree.

MAKE THIS REFER TO THE instructor-adjusted GPA in Lesson 12. Maybe use it to introduce **rank**.


Does adjusting the grades in this way make a difference? We can compare the `index` to the raw GPA, calculated in the conventional way.


ABOUT COMPARING THE THREE DIFFERENT FORMS OF ADJUSTMENT.

The data file containing the three forms of adjusted GPA is in ../../LSTtext/www/Three-gpas.rda" 

Introduce **sensitivity**, the idea that the result should not depend strongly on details which we don't think should be critical. Then introduce variance of the residuals from comparing each pair.

```{r label='250-Adjustment-bzqi1I', digits=1, message=FALSE, eval=FALSE}
# Convert to work with `Three_gpas`
Raw_gpa |>
  left_join(Adjusted_gpa) |>
  left_join(Model_adjusted_gpa) |>
  mutate(raw_vs_adj = rank(raw_gpa) - rank(grade_index),
         raw_vs_modeled = rank(raw_gpa) - rank(modeled_gpa),
         adj_vs_modeled = rank(grade_index) - rank(modeled_gpa)) |>
  select(contains("_vs_")) |> 
  pivot_longer(cols = contains("_vs_"), names_to = "comparison",
               values_to = "change_in_rank") |>
  summarize(var(change_in_rank), .by = comparison) |>
  kable()
```

This is, admittedly, a lot of wrangling. The result is that the two methods of adjustment agree with one another---a smaller variance of the change in rank---much more than the raw GPA agrees with either. This suggests that the adjustment is identifying a genuine pattern rather than merely randomly shifting things around.

:::
